### 2018-03-02 submission/manuscript.Rmd
flow update

Although we expected that the full OTU models would perform the best at classifying individuals with and without carcinomas, our observations suggest that both the full and significant OR taxa models performed equally well. These results suggest that lower level classification to species and strain may not add extra useful information with respect to prediction models. This has been suggested in previous literature where metagenomics did not perform better than 16S rRNA gene sequencing data at classifying individuals with normal colons and those with carcinomas [@Hannigan2017]. One possible reason as to why lower level classification may not result in better models is that the communities are patchy and higher level taxonomic information pools some of this patchiness allowing for better prediction models. There may also be a fair bit of data redundancy within models that utilize more of the community. An example of this redundancy would be when we trained models on one study and tested it on the other studies and the AUCs of the models created with the select OR taxa performed as well as full taxa models [Figure 6B].

Our observations also suggest that a small collection of taxa can classify diease as well as full OTU-based models but that these taxa individually perform quite poorly [Figure 3]. This result supports the contention that there might be redundancy of function even amongst the taxa included in the significant OR models. As an example multiple different microbes could be similarily stimulating the activation of inflammatory pathways and by doing so exacerbate disease progression. Multiple reports within the literature have found that different bacteria, such as *Escherichia coli* and *Fusobacterium nucleatum*, can worsen similarily worsen inflammation in mouse models of tumorigenesis [@ecoli_Arthur_2014; @arthur_intestinal_2012; @kostic_fusobacterium_2013]. Although the inflammatory taxa were patchy in their importance and presence across studies those that were not typically associated with inflammation were consistently important for every study [Figure 6]. The loss of these taxa (*Ruminococcus* and *Clostridium XI* in stool and *Dorea* and *Blautia* in unmatched tissue) is particularly interesting because many are commonly thought to be beneficial due to their involvement in production of short chain fatty acids [@Venkataraman2016; @Herrmann2018; @Reichardt2017].

The adenoma models as a whole performed poorly in classifying individuals with and without adenomas. This outcome is not inconsistent with what has been published previously [@hale_shifts_2017; @Sze2017]. However, the modeling results are at odds with results obtained in Baxter, *et al.* [@baxter_microbiota-based_2016]. There are some major differences between the models generated in this meta-analysis and what was used in this previous report. First, the prevoius report's models investigated the classification of lesions (individuals with adenoma or carcinoma) and not adenoma alone. The Baxter, *et al.* models also contained Fecal Immunoglobulin Test data while our meta-analysis models only contained 16S rRNA gene sequencing data. Although being able to classify individuals with adenomas is important, the classification of advanced adenomas is a more clinically meaningful diagnostic tool (i.e. those that are at high risk of progressing to a carcinoma). It is possible that we might have been able to detect differences in the bacterial community if advanced adenomas were separated from adenomas but that data was not available for the majority of studies analyzed. It is also possible that the initial changes to the bacterial community are focal to where the initial adenoma develops and would not be easily assessed with a fecal sample.

Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities. Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Sampling the microbiota directly associated with colon tissue may provide clearer answers but the colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with tumors. Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S3]. *Fusobacterium* was not identified consistently across studies and this could be due to both a small number of studies and a small sample size within these studies. Additionally, the majority of the colon tissue-based results were study specific with many of the top 10 taxa being present only in a single study. The presence of genera associated with contamination [@Salter_contamination_2014], within the top 10 most important variables for the genera and OTU models is worrying (e.g. *Novosphingobium*, *Acidobacteria Gp2*, *Sphingomonas*, etc.). The low bacterial biomass of tissue samples coupled with potential contamination and small sample sizes could explain why these results seem to be more sporadic than the stool results.

One important caveat to this study is that even though genera associated with certain species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp. *gallolyticus* were not identified, it does not necessarily mean that these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017]. There are reports that *Bacteroides fragilis*, positive for the enterotoxigenic gene, are found at specific locations along the colon but the samples we were able to use in this meta-analysis could not identify these types of differences [@Purcell2017]. Additionally, since we are limited in our aggregation of the data to the genus level, it is not possible to clearly delineate which species are contributing to overall disease progression. Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp. *gallolyticus*. As an example, the stool-based full community models consistently identified the genus *Bacteroides*, as well as OTUs that classified as *Bacteroides*, to be important model components across studies. This suggests that even though *Bacteroides* may not increase the OR of individuals having an adenoma or carcinoma and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC. Additionally, *Streptococcus gallolyticus* subsp. *gallolyticus* is a mouth-associated microbe, and the results from this study suggest that regardless of sample type, mouth-associated genera are commonly associated with an increased OR for individuals to have a carcinoma tumor.

Despite these limitations the findings that we present here would not be possible without performing a meta-analysis. These types of studies can be a useful tool in microbiota research because they can both validate existing research and make new discoveries by pooling many independent investigations together. Yet, it is still difficult to perform these studies because of inaccessible 16S sequencing data, missing or vague metadata (e.g. which samples are carcioma and which are not), varying sequence quality, and multiple small data sets. Better attention to these specific problems could help to increase the reproducibility and replicability of microbiota studies and make it easier to perform these crucial meta-analyses. Moving forward, meta-analyses will be important tools to help aggregate and find commonalities across studies when investigating the microbiota in the context of a specific disease and more are needed [@Sze2016; @Walters2014; @Finucane2014; @shah_leveraging_2017].

By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in carcinoma tumors. Although further validation of the biomarkers presented here need to be undertaken, the replciability of the AUC of a specific collection of taxa across multiple studies suggests a strong potential for the use of the microbiota as a risk stratification tool for individuals with carcinomas.

>>>>

Although we expected that the full OTU models would perform the best at classifying individuals with and without carcinomas, our observations suggest that both the full and significant OR taxa models performed equally well. These results suggest that lower level classification to species and strain may not add extra useful information with respect to prediction models. This has been suggested in previous literature where metagenomics did not perform better than 16S rRNA gene sequencing data at classifying individuals with normal colons and those with carcinomas [@Hannigan2017]. One possible reason as to why lower level classification may not result in better models is that the communities are patchy and higher level taxonomic information pools some of this patchiness, allowing for better prediction models. There may also be a fair bit of data redundancy within models that utilize more of the community. An example of this redundancy would be when we trained models on one study and tested it on the other studies and the AUCs of the models created with the select OR taxa performed as well as full taxa models [Figure 6B].

Our observations also suggest that a small collection of taxa can classify disease as well as full OTU-based models but that these taxa individually perform quite poorly [Figure 3]. This result supports the contention that there might be redundancy of function amongst the taxa included in the significant OR models. As an example, multiple different microbes could be similarily stimulating the activation of inflammatory pathways and by doing so exacerbate disease progression. Multiple reports within the literature have found that different bacteria, such as *Escherichia coli* and *Fusobacterium nucleatum*, can similarily worsen inflammation in mouse models of tumorigenesis [@ecoli_Arthur_2014; @arthur_intestinal_2012; @kostic_fusobacterium_2013]. Although the inflammatory taxa were patchy in their importance and presence across studies those that were not typically associated with inflammation were consistently important for every study [Figure 6]. The loss of these taxa (*Ruminococcus* and *Clostridium XI* in stool and *Dorea* and *Blautia* in unmatched tissue) is particularly interesting because many are commonly thought to be beneficial due to their involvement in production of short chain fatty acids [@Venkataraman2016; @Herrmann2018; @Reichardt2017].

The adenoma models as a whole performed poorly in classifying individuals with and without adenomas. This outcome is not inconsistent with what has been published previously [@hale_shifts_2017; @Sze2017]. However, the modeling results are at odds with results obtained in Baxter, *et al.* [@baxter_microbiota-based_2016]. There are some major differences between the models generated in this meta-analysis and what was used in Baxter, *et al*. First, the prevoius report's models investigated the classification of lesions (individuals with adenoma or carcinoma) and not adenoma alone. The Baxter, *et al.* models also contained Fecal Immunoglobulin Test data while our meta-analysis models only contained 16S rRNA gene sequencing data. Although being able to classify individuals with adenomas is important, the classification of advanced adenomas is a more clinically meaningful diagnostic tool (i.e. those that are at high risk of progressing to a carcinoma). It is possible that we might have been able to detect differences in the bacterial community if advanced adenomas were separated from adenomas but that data was not available for the majority of studies analyzed. Additionally, the initial changes to the bacterial community could be focal to where the initial adenoma develops and would not be easily assessed with a fecal sample.

Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- tissue associated microbial communities. The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with tumors. Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S3]. *Fusobacterium* was not consistently identified across studies and this could be due to both a small number of studies and a small sample size within these studies. The majority of the colon tissue-based results were also study specific with many of the top 10 taxa being present only in a single study. Additionally, the presence of genera associated with contamination [@Salter_contamination_2014], within the top 10 most important variables for the genera and OTU models is worrying (e.g. *Novosphingobium*, *Acidobacteria Gp2*, *Sphingomonas*, etc.). The low bacterial biomass of tissue samples coupled with potential contamination and small sample sizes could explain why these results seem to be more sporadic than the stool results.

One important caveat to this study is that even though genera associated with certain species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp. *gallolyticus* were not identified, it does not necessarily mean that these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017]. There are reports that *Bacteroides fragilis*, positive for the enterotoxigenic gene, are found at specific locations along the colon but the samples we were able to use in this meta-analysis could not identify these types of differences [@Purcell2017]. Additionally, since we are limited in our aggregation of the data to the genus level, it is not possible to clearly delineate which species are contributing to overall disease progression. Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp. *gallolyticus*. As an example, the stool-based full community models consistently identified the *Bacteroides* taxa to be important model components across studies. This suggests that even though *Bacteroides* may not increase the OR of individuals having an adenoma or carcinoma and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC.

Despite these limitations to the findings, meta-analyses are a useful tool in microbiota research because they can both validate existing research and make new discoveries by pooling many independent investigations together. Yet, it is still difficult to perform these studies because of inaccessible 16S sequencing data, missing or vague metadata (e.g. which samples are carcioma and which are not), varying sequence quality, and multiple small data sets. Better attention to these specific problems could help to increase the reproducibility and replicability of microbiota studies and make it easier to perform these crucial meta-analyses. Moving forward, meta-analyses will be important tools to help aggregate and find commonalities across studies when investigating the microbiota in the context of a specific disease and more are needed [@Sze2016; @Walters2014; @Finucane2014; @shah_leveraging_2017].

By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in carcinoma tumors. Although further validation of the biomarkers presented here need to be undertaken, the replicability of the AUC of a specific collection of taxa across multiple studies suggests a strong potential for the use of the microbiota as a risk stratification tool for individuals with carcinomas.
